---
title: "Data Analysis"
author: "Csaba Kertész"
date: '2021 01 09 '
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(tidytuesdayR)
library(funModeling)
library(Hmisc)
library(ggpubr)
library(scales)
library(RColorBrewer)
library(viridis)
library(car)
library(lm.beta)
library(psych)
```

# Income gap in Australia

The dataset was downloaded from data.gov.au. It contains data about income in different professions for both genders, published in 2008. In the following analysis I'm trying to explore the best and worst paying occupations for men and women, the wage gap between them, the possible association between yearly income, number of people in the same field and gender. 


```{r}
#importing dataset
wages <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-04-23/week4_australian_salary.csv")
wages_tb <- as_tibble(wages)


#checking for missing values
sapply(wages_tb,function(x) sum(is.na(x)))


#basic descriptives and plots
glimpse(wages_tb)
describe(wages_tb)
summary(wages_tb)
plot_num(wages_tb)

wages_tb %>% 
  ggplot(aes(x=gender, y=average_taxable_income, fill = gender)) +
  geom_boxplot() + geom_jitter(width=0.1,alpha=0.2)

#adding income range variable
wages_tb <- wages_tb %>%
mutate(income_rg = ifelse(wages_tb$average_taxable_income<summary(wages_tb$average_taxable_income)[2], "Low income", ifelse(wages_tb$average_taxable_income>summary(wages_tb$average_taxable_income)[5], "High income", "Average income")))


#summarizing variables by gender
describeBy(wages_tb, wages_tb$gender)


#plot 1 - Distribution of income among male and female workers
wages_tb%>%
  mutate(income_rg = fct_relevel(income_rg, 
            "Low income", "Average income", "High income")) %>%
ggplot(aes(x = gender, fill = income_rg)) +
    geom_bar(position = "fill") +
    scale_x_discrete(name = "") +
  scale_fill_viridis_d()+
    ylab(label="percentage (%)")+
    labs(fill = "")+
    theme_light() +
    coord_flip()+
    ggtitle("Distribution of income among genders") 

```
Missing values, outliers, basic descriptives

There were no missing values in the dataset. Two outliers were identified in the "average taxable income" col. and were decided to be kept for this part of the analysis.
The mean yearly taxable income in Australia was 65.678 (AUS), with a minimum of 13.307 and a maximum of 577.674.
Exploring income by gender, I found that men altogether earned more. Mean income was 75.152/56.109, with a minimum 16.490/13.307 and maximum of 577.674/355.844.  
Looking at plot 1, we see that while similar proportion of men and women work in average income jobs, a larger percentage of men work in the high income range, while a higher proportion of women have low income professions.
```{r}



#filtering male vs female occupations in order of income
wages_order_male <- wages_tb %>% 
      filter(gender == "Male") %>% 
      arrange(desc(average_taxable_income)) %>% 
      select(occupation, `Male Average Taxable Income` = average_taxable_income) 

wages_order_male <- wages_order_male %>%
  rename(male_income = 2)

wages_order_female <- wages_tb %>% 
      filter(gender == "Female") %>% 
      arrange(desc(average_taxable_income)) %>% 
      select(occupation, `Female Average Taxable Income` = average_taxable_income)

wages_order_female <- wages_order_female %>%
  rename(female_income = 2)


#pairing by same occupation, sorted by male income
occ_by_male  <- wages_order_male %>% 
      left_join(., wages_order_female , by = "occupation") %>% 
      arrange(desc(male_income)) 

#adding income range variable
occ_by_male <- occ_by_male %>%
  mutate(income_rg = ifelse(occ_by_male$male_income<summary(occ_by_male$male_income)[2], "Low income", ifelse(occ_by_male$male_income>summary(occ_by_male$male_income)[5], "High income", "Average income")))

#calculating relative income gap
occ_by_male <-  occ_by_male %>%
  mutate(gap = (male_income - female_income)/male_income)

occ_by_male <- occ_by_male %>%
  mutate(gap_percent = gap * 100)

occ_by_male %>%
  summary()

female_adv <-  occ_by_male %>%
  filter(female_income > male_income) 

female_adv %>%
  summary()
  

#plot2 - Comparing female and male income in the same professions
occ_by_male%>%
  ggplot(aes(x = male_income , y = female_income, color=income_rg)) + 
  geom_point() +
  scale_y_continuous(limits = c(0, 500000), 
                     name = "Male income (AUS)", labels = function(x) format(x, scientific = FALSE))+
  scale_x_continuous(limits = c(0, 500000),
                     name = "Female income (AUS)",labels = function(x) format(x, scientific = FALSE))+
    theme_light() +
    geom_abline()+
    scale_color_viridis_d(name = "")+
    ggtitle("Female and male yearly income in the same professions") 


```
The GAP

When we compare the income of men and women in the same professions, we see that men earn 19.74% more in average. The maximum relative wage gap is 83.4% (for cricketers) in favor of men. Interestingly we found that there is an extreme relative difference (233.1% for mountaineers), in which case there is a female advantage. 

Looking at plot2, we see that there are very few occupations (81, which is only 7,3%) in which women earn more, most of which are in the "low income" range (56, 69%).


```{r}


 #identifying outliers in gap variable 
out <- occ_by_male %>%
  arrange(gap_percent)


#removing 5 rows containing outliers
occ_by_male <-  occ_by_male[-c(1032), ] 
occ_by_male <-  occ_by_male[-c(821), ] 
occ_by_male <-  occ_by_male[-c(406), ] 
occ_by_male <-  occ_by_male[-c(961), ] 


#checking normality for variables - all 3 variables differ from normal distribution
shapiro.test(occ_by_male$male_income)
shapiro.test(occ_by_male$female_income)
shapiro.test(occ_by_male$gap)


cor.test(occ_by_male$male_income, occ_by_male$gap, method = c("s")) 

#plot3 - Correlation between yearly income and wage gap in 3 income categories
occ_by_male %>%
   mutate(income_rg = fct_relevel(income_rg, 
            "Low income", "Average income", "High income")) %>%
ggplot(aes(x=male_income, y=gap_percent, color = income_rg)) +
  theme_light() +
  geom_point(size =0.5) +
scale_y_continuous(name = "Wage gap in percent", labels = function(x) format(x, scientific = FALSE))+
  scale_x_continuous(name = "Yearly male income (AUS)",labels = function(x) format(x, scientific = FALSE))+
  scale_color_viridis_d(name = "")+
  geom_smooth(method=lm)+
  stat_cor(method = "spearman")+
  ggtitle("Correlation between yearly income and wage gap")


```
The correlation between wage gap and income 

Our hypothesis was that the relative wage gap widens as income grows. To test our hypothesis we calculated Spearman correlation  for the two variables and found that there is in fact a positive association (rho =0.4978624 , p = 2.2e-16) between them. The higher the average income, the greater the gap becomes.

Plot3 also shows that the correlation between income and the wage gap persists in each income category with a weak association in the average income range, and moderate (R = 0.42) in the "high" and "low" categories.
```{r}


#BUILDING MODELS to predict income with gender and individuals variables

#testing normality of target variable - normality assumption not met
shapiro.test(wages_tb$average_taxable_income)


#normalizing target variable with z-transformation
normalize = function(x){
  return(((x - min(x, na.rm = TRUE))/(max(x, na.rm = TRUE))))
}

     
wages_tb <- wages_tb %>%
  mutate(income_norm = normalize(average_taxable_income))


#Building first linear model
model_1 <- lm(income_norm ~ gender + individuals, data = wages_tb)
summary(model_1)


#model diagnostics and outliers
model_1 %>%
  plot(5)

model_1 %>%
  plot(2)

model_1 %>%
  plot(1)

model_1 %>%
  plot(3)

model_1 %>%
  vif()

#removing outliers
wages_tb<- wages_tb[-c(1550, 1598, 482, 548, 1682, 1202),]

#Building second model
model_2 <- lm(income_norm ~ gender, data = wages_tb)
summary(model_2)

#model diagnostics and outliers
model_2 %>%
  plot(5)

model_2 %>%
  plot(2)

model_2 %>%
  plot(1)

model_2 %>%
  plot(3)

#removing outliers
wages_tb <- wages_tb[-c(1609),]

model_1 <- lm(income_norm ~ gender + individuals, data = wages_tb)
model_2 <- lm(income_norm ~ gender, data = wages_tb)

model_2 %>%
lm.beta()

model_2 %>%
coef()

model_2 %>%
confint(level = 0.95)

anova(model_1, model_2)
```
Building a model for predicting income

My first model for predicting income included "gender" and "individuals" variables. 
After removing influential outliers, the model still didn't meet the normality assumption. The model accounted for 3,34% of the total variance.
Adjusted R-squared:  0.03436 F-statistic: 40.07 on 2 and 2194 DF,  p-value: < 2.2e-16

"Individuals" variable was dropped from the second model, leaving only "gender". This model explained 3,35% of the variance of the income variable. Just as with the first model, this one also violated the normality assumption.
Adjusted R-squared:  0.03358 F-statistic:  77.1 on 1 and 2189 DF,  p-value: < 2.2e-16
Equation: y = 0.030760 + 0.003507 * x1

ANOVA shows that the two models do not differ significantly.

SUMMARY
The analysis shows that although gender plays an important role, it is not sufficient in itself to predict income. We hypothesize that the field in which an individual is working has a robust effect on wages. Unfortunately the dataset doesn't contain a categorical variable for grouping the many kinds of occupations. We created a subset dataset with a "field" variable containing 6 values that we could extract from the "occupations" variable using string matching. This second analysis is only an attempt to get a sense of how robust an effect the field of work has on income (by common sense I hypothesize that a strong one). 


```{r}

#identifying some fields among occupations for later analyses by string match
scientist=wages_tb[grep("scien",wages_tb$occupation),] 
engineer=wages_tb[grep("engineer",wages_tb$occupation),]
worker=wages_tb[grep("worker",wages_tb$occupation),]
nurse=wages_tb[grep("nurse",wages_tb$occupation),]
surgeon=wages_tb[grep("surgeon",wages_tb$occupation),]
cook=wages_tb[grep("cook",wages_tb$occupation),]

scientific <- scientist %>%
  mutate(field = "science")

engineer <- engineer %>%
  mutate(field = "engineer")

worker <- worker %>%
  mutate(field = "worker")

nurse <- nurse %>%
  mutate(field = "nurse")

surgeon <- surgeon %>%
  mutate(field = "surgeon")

cook <- cook %>%
  mutate(field = "cook")

#creating subset table with "field" variable
wages_tb_field <-  rbind(cook, worker, surgeon, nurse, scientific, engineer)


#descriptives by field
describeBy(wages_tb_field, wages_tb_field$field)


#plot4 - The income gap in different fields 
wages_tb_field %>%
ggplot(aes(x = field, y = average_taxable_income, fill = gender)) + geom_col(position = position_dodge(width = 0.8), width = 0.7)+ 
  scale_fill_viridis_d() +
   theme(axis.text.x = element_text(size = 10, angle = 45)) +
  scale_y_continuous(name = "Average taxable icnome (AUS)", labels = function(x) format(x, scientific = FALSE))+
  scale_x_discrete(name = "",labels = function(x) format(x, scientific = FALSE))+
  theme_light() +
  ggtitle("Average yearly income by fields and gender") 



#Building linear model for the subset dataset
model_3 <- lm(income_norm ~ gender + field, data = wages_tb_field)
summary(model_3)

#Model diagnostics and outliers
model_3 %>%
  plot(5)

model_3 %>%
  plot(2)

model_3 %>%
  plot(1)

model_3 %>%
  plot(3)

#removing outliers
wages_tb_field <- wages_tb_field[-c(131, 132, 134, 135, 141),]

model_3 <- lm(income_norm ~ gender + field, data = wages_tb_field)
summary(model_3)

model_3 %>%
  vif()

model_3 %>%
lm.beta()

model_3 %>%
coef()

model_3 %>%
confint(level = 0.95)


```
SUMMARY - subset dataset

The third, exploratory model included "gender" and "field" as variables to predict "income" After removing outliers the model met the normality and linearity assumption, but violated homoscedasticty. Multicollinearity was acceptable as VIF for both variables was under 5. This model accounted for 72% of variance of the "income" variable.

Adjusted R-squared:  0.7206 
F-statistic: 131.7 on 6 and 298 DF,  p-value: < 2.2e-16
Equation_ y = 0.008805 + 0.004927 * x1 + 0.014244 * x2 + 0.015160 * x3 + 0.016179 * x4 + 0.020399  * x5 + 0.014148 * x6

Exploring each variable's contribution by their standard coefficients (beta), the "field" variable's values showed strong influence. Variables in order of beta value were: surgeon - 0,85; engineer - 0,56; science - 0,23; gender (male) - 0,21; worker - 0,17.


CONCLUSION
All variables showed a significant contribution  to the total variance, with "field" far overweighing "gender". Plot4 also shows that the gap between male and female income widens in higher income fields, confirming the earlier conclusion drawn from the whole dataset. 

As this was only an exploratory model with a fraction of the whole dataset, there can be no far reaching conclusions drawn from this last analysis, but the results stress the importance of the dataset be updated by a categorical variable for the different professions if we plan to build better models.




