---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(scales)
library(viridis)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
tuesdata <- tidytuesdayR::tt_load(2020, week = 39)
expeditions <- tuesdata$expeditions

expeditions_clean <- expeditions %>%
  mutate(peaks_clean = fct_lump_n(expeditions$peak_name, n = 15))


expeditions_clean <- expeditions_clean %>%
  mutate(peaks_clean = fct_lump_n(expeditions$peak_name, n = 15)) #level összevon,


 expeditions_clean <- filter(expeditions_clean, peaks_clean != "Other")%>%
   droplevels()  # other-drop

 expeditions_clean <- expeditions_clean %>%
group_by(peaks_clean) %>%
  mutate(freq=n())%>%
  arrange(desc(freq)) #freq változó
  
expeditions_clean$peaks_clean <-fct_reorder(expeditions_clean$peaks_clean, .x=expeditions_clean$freq, .desc = FALSE) #freq alapján sorba rak


  
ggplot(expeditions_clean, aes(x = peaks_clean, fill=season)) + 
    geom_bar()  + 
    coord_flip() +
  theme_light() +
  theme(legend.position = "bottom")+
  scale_fill_viridis(discrete = TRUE)+
  scale_x_discrete(name="") +
  ggtitle("The 15 most popular peaks stacked by season of expedition")+
  scale_y_continuous(name="Number of expeditions")
  
  
```

## Task 2: PhDs awarded

The group_by(broad_field, year) %>% 
  summarize(n_phds = sum(n_phds)) dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
tuesdata_2 <- tt_load("2019-02-19")
phd <- tuesdata_2$phd_by_field

phd_field_year <- phd %>%
    group_by(broad_field, year) %>% 
    summarize(n_phds = sum(n_phds, na.rm=TRUE)) 

ggplot(data=phd_field_year, aes(x=year, y=n_phds, colour=broad_field)) +
  theme_light() +
  labs(color = "Broad field")+
  geom_line(size =1.2) +
   scale_color_brewer(palette = "Dark2")+
   scale_x_continuous(name="", breaks=c(2008,2010,2012,2014,2016)) +
  scale_y_continuous(name="") +
    ggtitle("Number of awarded PhD-s in the US by year ")
  
 
 
  
  

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
tuesdata_2 <- tt_load("2019-11-05")
commute <- tuesdata_2$commute

commute_group <- commute %>%
     group_by(state_abb, mode, state_region) %>% 
     summarize(n = sum(n, na.rm=TRUE))

walk <- commute_group %>%
    filter(mode == "Walk")


bike <- commute_group %>%
    filter(mode == "Bike") 

commute_tbl <- tibble(walk$n, bike$n, walk$state_abb, walk$mode)

 ggplot(commute_tbl, aes(x=walk$n, y=bike$n, colour=walk$state_region)) + 
  geom_point() +
   guides(color = guide_legend(override.aes = list(size = 1)))+
   theme_light()+
   theme(legend.title = element_blank())+
      scale_x_continuous(trans = "log10", labels = scales::number_format(               decimal.mark = ','), name = "")+
   scale_y_continuous(trans = "log10", labels = scales::number_format(               decimal.mark = ','), name = "") +
   geom_text(aes(label = walk$state_abb), color="black")+
   ggtitle("Title number of people walking vs. biking to work in each US state")
   
     
 
   

```
